---
title: "500 HW3"
author: "Diana Liang"
date: "10/9/2019"
output:
  word_document: default
  html_document: default
---

```{r setup}

library(faraway)
data("sat")

#A1: sat lm model
sat_model = lm(total ~ takers + ratio + salary, data = sat)
summary(sat_model)

#A2: salary t test
salary_tvalue = 2.541
salary_pvalue = 1-pt(2.541, df = 46)
cat("Salary T-Test Statistic:", salary_tvalue, 
    "\n", "Salary P-Value:",salary_pvalue)

#A3: ratio t test
ratio_tvalue = -2.187
ratio_pvalue = 2 *(pt(ratio_tvalue, df=46))
cat("Ratio T-Test Statistic:", ratio_tvalue, "\n",
    "Ratio P-Value:", ratio_pvalue)

#A4: sat f test
cat("\n")
cat("SAT data F-Test Statistic: 71.72", 
    "\n", "SAT data P-Value: 0")

#A5: salary 95 and 99 ci
salary_conf95 = confint(sat_model, "salary", level = 0.95)
salary_conf95
salary_conf99 = confint(sat_model, "salary", level = 0.99)
salary_conf99

#A6: joint 95 ci
library(ellipse)
plot(ellipse(sat_model, c('ratio', 'salary')), main = "95% Joint Confidence Region for Ratio and Salary", 
     type = "l")
points(0,0, pch=1)
#A6: F test for a pair
ho_6 = lm(total ~ takers, sat)
hoa_6 = lm(total ~takers + ratio + salary, sat)
anova(ho_6, hoa_6)

#A7: sat lm now with expend
sat_model7 = lm(total ~ takers + ratio + salary + expend, sat)
summary(sat_model7)

#A8: f test for pair salary, expend, ratio
ho_8 = lm(total ~ takers, sat)
hoa_8 = lm(total ~ takers + ratio + salary + expend, sat)
anova(ho_8, hoa_8)
```

```{r}
data(teengamb)
teengamb$sex = factor(teengamb$sex)
levels(teengamb$sex) = c("Male", "Female")
gamb_model = lm(gamble ~ sex + status + income + verbal, teengamb)
summary(gamb_model)
gamb_model_s = lm(sqrt(gamble) ~ sex + status + income + verbal, teengamb)
summary(gamb_model_s)

#B1: constant variance
plot(gamb_model$fitted, gamb_model$residuals, main = "Fitted Gamble Expenditure vs Residuals",
     xlab = "Fitted", ylab = "Residuals")
summary(lm(abs(gamb_model$residuals) ~ gamb_model$fitted))
plot(gamb_model_s$fitted, gamb_model_s$residuals,
     main = "Fitted Sqrt Gamble Expenditures vs Residuals", xlab = "Fitted", ylab = "Residuals")
summary(lm(abs(gamb_model_s$residuals) ~ gamb_model_s$fitted))

#B2: normality
qqnorm(gamb_model_s$residuals, main= "Residual Normality", ylab = "Residuals")
qqline(gamb_model_s$residuals)
shapiro.test(gamb_model_s$residuals)

#B3: leverage points
halfnorm(lm.influence(gamb_model_s)$hat, nlab = 4, main = "High Leverage Points", ylab = "Leverages")
lm.influence(gamb_model_s)$hat[lm.influence(gamb_model_s)$hat>(2*(4+1)/47)]

#B4: outliers
outlie = rstudent(gamb_model_s)
m = max(abs(outlie))
which(outlie == m)
p_val=2*(1-pt(m, df = 47-5-1))
Bonf_alpha=0.05/47
cat("Point with Max Studentized RS:", which(outlie == m),
    "\n", "P-value for Possible Outlier:", p_val,
    "\n", "Corrected Significance Level:", Bonf_alpha)

#B5: influential
cook = cooks.distance(gamb_model_s)
halfnorm(cook, nlab = 4, main = "Influential Points", ylab = "Cook's Distance")
gamb_mod_24 = lm(sqrt(gamble) ~ sex + status + income + verbal, data = teengamb, subset = (cook < max(cook)))
summary(gamb_mod_24)

gamb_model_s_inf = lm.influence(gamb_model_s)

plot(gamb_model_s_inf$coef[,3], gamb_model_s_inf$coef[,5],
     main = "Change in betas Status and Verbal", xlab = "Change in beta Status", ylab = "Change in beta Verbal")
identify(gamb_model_s_inf$coef[,3], gamb_model_s_inf$coef[,5])

plot(gamb_model_s_inf$coef[,3], gamb_model_s_inf$coef[,4],
     main = "Change in betas Status and Income", xlab = "Change in beta Status", ylab = "Change in beta Income")
identify(gamb_model_s_inf$coef[,3], gamb_model_s_inf$coef[,4])

plot(gamb_model_s_inf$coef[,4], gamb_model_s_inf$coef[,5],
     main = "Change in betas Income and Verbal", xlab = "Change in beta Income", ylab = "Change in beta Verbal")
identify(gamb_model_s_inf$coef[,4], gamb_model_s_inf$coef[,5])

# B6: structure of relationship
# Partial regression plot
delta <- residuals(lm(sqrt(gamble) ~ sex + income + verbal, teengamb))
gamma <- residuals(lm(status ~ sex + income + verbal, teengamb))
plot(gamma, delta, main = "Partial Regression for Status", xlab = "Status Residuals", ylab = "Sqrt Gamble Residuals")
temp <- lm(delta~gamma)
abline(reg = temp)
identify(gamma,delta)

delta <- residuals(lm(sqrt(gamble) ~ sex + status + verbal, teengamb))
gamma <- residuals(lm(income ~ sex + status + verbal, teengamb))
plot(gamma, delta, main = "Partial Regression for Income", xlab = "Income Residuals", ylab = "Sqrt Gamble Residuals")
temp <- lm(delta~gamma)
abline(reg = temp)
identify(gamma,delta)

delta <- residuals(lm(sqrt(gamble) ~ sex + income + status, teengamb))
gamma <- residuals(lm(verbal ~ sex + income + status, teengamb))
plot(gamma, delta, main = "Partial Regression for Verbal", xlab = "Verbal Residuals", ylab = "Sqrt Gamble Residuals")
temp <- lm(delta~gamma)
abline(reg = temp)
identify(gamma,delta)


# Partial residual plot
plot(teengamb$status, gamb_model_s$residuals +
         coef(gamb_model_s)['status']*teengamb$status, main = "Partial Residual for Status",
     xlab="Status", ylab= "Sqrt Gamble (adjusted)")
abline(a=0, b=coef(gamb_model_s)['status'])

plot(teengamb$income, gamb_model_s$residuals +
         coef(gamb_model_s)['income']*teengamb$income, main = "Partial Residual for Income",
     xlab="Income", ylab= "Sqrt Gamble (adjusted)")
abline(a=0, b=coef(gamb_model_s)['income'])

plot(teengamb$verbal, gamb_model_s$residuals +
         coef(gamb_model_s)['verbal']*teengamb$verbal, main = "Partial Residual for Verbal",
     xlab="Verbal", ylab= "Sqrt Gamble (adjusted)")
abline(a=0, b=coef(gamb_model_s)['verbal'])
