---
title: "500 HW#4"
author: "Diana Liang"
date: "10/27/2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(faraway)
data(longley)
```

Full Model
```{r, echo = FALSE}
full <- lm(Employed ~ GNP.deflator + GNP + Unemployed + Armed.Forces + Population + Year, longley)
summary(full)

#scaled <- data.frame(scale(longley))
#df <- lm(Employed ~ GNP.deflator + GNP + Unemployed + Armed.Forces + #Population + Year, scaled)
#summary(df)
```

Condition Numbers
```{r, echo = FALSE}
X <- model.matrix(full)[,-1]
e <- eigen(t(X) %*% X)
#e$val
round(sqrt(e$val[1]/e$val), 3)

#X <- model.matrix(df)[,-1] # scaled; all except intercept
#e <- eigen(t(X) %*% X)
#e$val
#round(sqrt(e$val[1]/e$val), 3)
```

Varation in predictors is 5000 times bigger between the largest and smallest variation. This means that some of the eigenvalues for the (X transpose)*X are, relatively, very small and that some of the predictor variables are collinear.

Correlation btwn predictors
```{r, echo = FALSE}
round(cor(longley)[1:6, 1:6], 2)
```

GNP deflator, GNP, Population and Year predictor variables have a correlation of nearly 1, meaning that they are most likely collinear. These variables should be condensed down to one predictor variable.


Variance inflation factors
```{r, echo = FALSE}
round(vif(X), 3)
```

GNP.deflator, GNP, Population and Year predictor variables had intensely greater variance inflation factors compared to Unemployed and Armed.Forces, meaning that those 4 variables are very highly correlated. 


Reduced Model
```{r, echo = FALSE}
reduced <- lm(Employed ~ GNP.deflator + Unemployed + Armed.Forces, longley)
summary(reduced)
```

The full model showed that the predictor variables Unemployed, Armed.Forces, and Year were significant, with a significance level of 0.05, in modelling the response variable Employed. The reduced model now only has predictor variables GNP.deflator and Unemployed as significant in modeeling Employed.


Reduced Model Condition Numbers
```{r, echo = FALSE}
X_red <- model.matrix(reduced)[,-1]
e_red <- eigen(t(X_red) %*% X_red)
#e_red$val
round(sqrt(e_red$val[1]/e_red$val), 3)
```

The reduced model shows relatively low eigenvalues for the remaining predictor variables, so there is little evidence of collinearity between them.

Reduced Model Variance Inflation Factors
```{r, echo = FALSE}
round(vif(X_red), 3)
```

The variance inflation factors are very similar and quite, which confirms there is little correlation between the remaining predictor variables.